Prediction about 2024 U.S. Presidential Election Outcome Using Linear Modeling

### 2024-10-25 paper/paper.qmd
Updated model section of the draft

library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(rstanarm)
library(splines)
library(ggplot2)
```

@@

analysis_data <- 
  read_csv(file = here::here("data/02-analysis_data/poll_analysis_data.csv"))

just_trump_data <- 
  analysis_data |> filter(candidate_name == "Donald Trump" | 
                          candidate_name == "Donald Trump Jr.") |>
  mutate(num_trump = round((pct / 100) * sample_size, 0))

just_trump_data |>
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth() +
  theme_classic() +
  labs(y = "Trump percent", x = "Sample size")
```

Color by pollster

# Color by pollster
just_trump_data |>
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point(aes(color = pollster)) +
  geom_smooth() +
  theme(legend.position = "bottom") +
  labs(ylab = "Trump percent", xlab = "Sample size")
```

Facet by pollster

# Facet by pollster
just_trump_data |>
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point() +
  geom_smooth() +
  facet_wrap(vars(pollster)) +
  theme_classic() +
  labs(ylab = "Trump percent", xlab = "Sample size")
```

Color by transparency

# Facet by pollster
just_trump_data |>
  ggplot(aes(x = sample_size, y = pct)) +
  geom_point(aes(color = factor(transparency_score))) +
  geom_smooth() +
  theme(legend.position = "bottom") +
  labs(ylab = "Trump percent", xlab = "Sample size")
```

@@

model_size <-
  readRDS(file = here::here("models/trump_model_size.rds"))

model_size_pollster <-
  readRDS(file = here::here("models/trump_model_size_pollster.rds"))
```

@@

modelsummary::modelsummary(
  list(
    "Model 1" = model_size,
    "Model 2" = model_size_pollster
  ),
  fmt = 2
)
```

>>>>

library(tidyverse)
library(janitor)
library(lubridate)
library(broom)
library(modelsummary)
library(rstanarm)
library(splines)
library(ggplot2)
library(arrow)
```

@@

# Read data
trump_national_data <-
  read_parquet(
    file = here::here("data/02-analysis_data/trump_national_data.parquet")
    )
trump_state_data <-
  read_parquet(
    file = here::here("data/02-analysis_data/trump_state_data.parquet")
    )

```

@@

spline_model_national <-
  readRDS(file = here::here("models/model_national.rds"))

spline_model_state <-
  readRDS(file = here::here("models/model_state.rds"))
```

@@

modelsummary::modelsummary(
  list(
    "Model 1" = spline_model_national,
    "Model 2" = spline_model_state
  ),
  fmt = 2
)
```

Posterior predictive checks for national spline model:

pp_check(spline_model_national)
```

Posterior predictive checks for state spline model:

pp_check(spline_model_state)
```

Predict posterior draws and spline fit for votes nationally

# Predict and plot
# Create new data for prediction
new_data_national <- data.frame(
  end_date_num = seq(
    min(trump_national_data$end_date_num),
    max(trump_national_data$end_date_num),
    length.out = 100
  )
)

# Predict posterior draws
posterior_preds_national <- posterior_predict(spline_model_national, 
                                     newdata = new_data_national)

# Summarize predictions
pred_summary_national <- new_data_national |>
  mutate(
    pred_mean = colMeans(posterior_preds_national),
    pred_lower = apply(posterior_preds_national, 2, quantile, probs = 0.025),
    pred_upper = apply(posterior_preds_national, 2, quantile, probs = 0.975)
  )

# Plot the spline fit
ggplot(trump_national_data, aes(x = end_date_num, y = pct)) +
  geom_point() +
  geom_line(
    data = pred_summary_national,
    aes(x = end_date_num, y = pred_mean),
    color = "blue",
    inherit.aes = FALSE
  ) +
  geom_ribbon(
    data = pred_summary_national,
    aes(x = end_date_num, ymin = pred_lower, ymax = pred_upper),
    alpha = 0.2,
    inherit.aes = FALSE
  ) +
  labs(
    x = "Number of Days Since Half a Year Before Election",
    y = "Vote Percentage for Trump",
    title = "Poll Percentage of Trump over Time with Spline Fit"
  ) +
  theme_minimal()
```

Predict posterior draws and spline fit for votes in each state

# Change 'pollster' and 'state' to factor variables
trump_state_data <- trump_state_data |>
  mutate(
    pollster = factor(pollster),
    state = factor(state)
  )

# Predict and plot
# Create new data for prediction
new_data_state <- expand.grid(
  end_date_num = seq(
    min(trump_state_data$end_date_num),
    max(trump_state_data$end_date_num),
    length.out = 100
  ),
  state = levels(trump_state_data$state)  # Include all states
)

# Predict posterior draws
posterior_preds_state <- posterior_predict(spline_model_state, 
                                              newdata = new_data_state)

# Summarize predictions
pred_summary_state <- new_data_state |>
  mutate(
    pred_mean = colMeans(posterior_preds_state),
    pred_lower = apply(posterior_preds_state, 2, quantile, probs = 0.025),
    pred_upper = apply(posterior_preds_state, 2, quantile, probs = 0.975)
  )

# Plot the spline fit
ggplot(trump_state_data, aes(x = end_date_num, y = pct, color = state)) +
  geom_point() +
  geom_line(data = pred_summary_state, 
            aes(x = end_date_num, y = pred_mean, group = state), 
            color = "blue") +
  labs(
    x = "Number of Days Since Half a Year Before Election",
    y = "Vote Percentage for Trump",
    title = "Poll Percentage of Trump over Time with Spline Fit"
  ) +
  theme_minimal() +
  facet_wrap(~ state, scales = "free_y") +  # Faceting by state
  theme(
    strip.text = element_text(size = 7), # Title size for each facet
    axis.text = element_text(size = 5), # Axis text size
    legend.text = element_text(size = 5),  # Adjust legend text size
    legend.title = element_text(size = 5),  # Adjust legend title size
    legend.key.size = unit(0.5, "cm"),  # Adjust size of legend keys
    legend.position = "right"  # Position legend to the right
  )
```
